# STAT 540: Seminar 8
# Methylation analysis
# Date: Wed Mar 5
# By: Jessica Pilsworth
========================================================
Load libraries
```{r}
library(GEOquery)
library(wateRmelon)
library(IlluminaHumanMethylation450k.db)
```

# Explore 450k methylation array data

Load data
```{r}
if (file.exists("methyl_ALL.Rdata")) {
    # if previously downloaded
    load("methyl_ALL.Rdata")
} else {
    # if downloading for the first time
    GSE39141 <- getGEO("GSE39141")
    show(GSE39141)  ## 33 samples (29 ALL and 4 healthy B cells)
    GSE42865 <- getGEO("GSE42865")  # took ~2 mins for JB
    show(GSE42865)  ## 16 samples (9 healthy cells B cells and 7 other cells)

    # Extract expression matrices (turn into data frames at once)
    ALL.dat <- as.data.frame(exprs(GSE39141[[1]]))
    CTRL.dat <- as.data.frame(exprs(GSE42865[[1]]))

    # Obtain the meta-data for the samples and rename them perhaps?
    ALL.meta <- pData(phenoData(GSE39141[[1]]))
    CTRL.meta <- pData(phenoData(GSE42865[[1]]))

    # create some labels
    ALL.meta$Group <- c(rep("ALL", 29), rep("HBC", 4))
    ## ALL: Case; HBC: Healthy B Cells

    # Subset both meta-data and data for control (healthy) donors
    CTRL.meta <- droplevels(subset(CTRL.meta, grepl("Healthy donor", characteristics_ch1.1)))
    CTRL.dat <- subset(CTRL.dat, select = as.character(CTRL.meta$geo_accession))

    # Rename variables
    names(ALL.dat) <- paste(ALL.meta$Group, gsub("GSM", "", names(ALL.dat)), 
        sep = "_")
    names(CTRL.dat) <- paste("HBC", gsub("GSM", "", names(CTRL.dat)), sep = "_")

    # save the data to avoid future re-downloading
    save(ALL.dat, CTRL.dat, ALL.meta, CTRL.meta, file = "methyl_ALL.Rdata")
}
```

Load library
```{r}
library(ggplot2)
```

To calculate the average Beta values for the probes in the two datasets 
```{r}
pmDat <- c(rowMeans(ALL.dat, na.rm = T), rowMeans(CTRL.dat, na.rm = T)) 
```

To create data frame for the density plot
```{r}
probe.meansDat <- data.frame(Beta = pmDat,
                             Dataset = rep(c('ALL', 'CTRL'), each = nrow(ALL.dat)))
```

To create a density plot of the probe mean beta values 
```{r fig.width=7, fig.height=6}
(probe.meansPlot <- ggplot(data = probe.meansDat, aes(x = Beta, col = Dataset)) +
    geom_density() + ggtitle("Density Plot of Mean Beta Values") + 
    xlab("Beta") + ylab("Density"))
```

# Normalization

To combine data from two experiments into one matrix, each column represents beta values of one sample
```{r}
beta.matrix <- as.matrix(cbind(ALL.dat, CTRL.dat))
str(beta.matrix, max.level = 0)
```

To perform quantile normalization
To calculate normalized betas from Illumina 450K methylation arrays
```{r}
system.time(beta.norm <- betaqn(beta.matrix))
```

To calculate row means for normalized data
```{r}
probe.meansDatNorm <- c(rowMeans(beta.norm[, 1:ncol(ALL.dat)], na.rm = TRUE),
                    rowMeans(beta.norm[, ncol(ALL.dat):ncol(beta.norm)], na.rm = TRUE)) 
```

To create a normalized plot and add a level
```{r}
probe.meansNormPlot <- rbind(data.frame(probe.meansDat, Norm = "Before"),
                  data.frame(Beta = pmDat,
                   Dataset = rep(c('ALL', 'CTRL'), each = nrow(ALL.dat)),
                   Norm = "After"))
probe.meansNormPlot$Norm <- factor(probe.meansNormPlot$Norm, levels = c("Before", "After"))
```

To create a density plot of the probe mean beta values before and after normalization
```{r fig.width=7, fig.height=6}
(probe.meansNorm <- ggplot(data = probe.meansNormPlot, aes(x = Beta, col = Dataset)) +
   geom_density() + facet_grid(Norm ~ .) + 
   ggtitle("Density Plot of Mean Beta Values Before and After Normalization") + 
   xlab("Beta") + ylab("Density"))
```

# M values

To apply a logit transformation on the data to convert it to a continuous variable that spans (−∞,∞) i.e. compute the M value
```{r}
M.norm <- beta2m(beta.norm)
```

# CpG Islands

To extract probe ID to CpG islands association
```{r}
cginame <- as.data.frame(IlluminaHumanMethylation450kCPGINAME)
names(cginame) <- c("Probe_ID", "cginame")
rownames(cginame) <- cginame$Probe_ID
length(levels(factor(cginame$cginame))) 
```

To restrict probes to those within CGIs
```{r}
beta.inCGI <- beta.norm[cginame$Probe_ID, ]
M.inCGI <- M.norm[cginame$Probe_ID, ]
nrow(M.inCGI) 
```

To aggregate probes to CGIs
```{r}
beta.CGI <- aggregate(beta.inCGI, by = list(cginame$cginame), mean, na.rm = T)
rownames(beta.CGI) <- beta.CGI[, "Group.1"]
beta.CGI <- subset(beta.CGI, select = -Group.1)
str(beta.CGI, max.level = 0)

M.CGI <- aggregate(M.inCGI, by = list(cginame$cginame), mean, na.rm = T)
rownames(M.CGI) <- M.CGI[, "Group.1"]
M.CGI <- subset(M.CGI, select = -Group.1)
str(M.CGI, max.level = 0)
```

To reshape the data
```{r}
library(reshape2)
M.CGI.tall <- melt(t(M.CGI), value.name = "M", varnames = c("Sample", "CGI"))
M.CGI.tall$Group <- substring(M.CGI.tall$Sample, 1, 3)
```

To create of bloxplot of CGI M values
```{r}
(M.boxplot <- ggplot(data = M.CGI.tall, aes(Sample, M, colour= Group)) + 
   geom_boxplot() + ggtitle("Bloxplot of CGI M values") + 
   xlab("Samples") + ylab("M Values") + 
     scale_x_discrete(labels = NULL))
```


# Differential methylation analysis with limma

```{r}
library(limma)
design <- data.frame(Group = relevel(factor(gsub("_[0-9]+", "", colnames(M.CGI))), 
                                     ref = "HBC"), row.names = colnames(M.CGI))
str(design)
(DesMat <- model.matrix(~Group, design))
DMRfit <- lmFit(M.CGI, DesMat)
DMRfitEb <- eBayes(DMRfit)
cutoff <- 0.01
DMR <- topTable(DMRfitEb, coef = "GroupALL", number = Inf, p.value = cutoff)
head(DMR) # top hits
```

To create plots to check these hits
```{r}
library(gplots)
DMR100 <- topTable(DMRfitEb, coef = "GroupALL", number = 100)
DMR.CGI <- t(as.matrix(subset(beta.CGI, rownames(beta.CGI) %in% rownames(DMR100))))
str(DMR.CGI, max.level = 0)
```

To create heatmap of beta values of top 100 hits
```{r fig.width=7, fig.height=6}
col <- c(rep("darkgoldenrod1", times = nrow(DMR.CGI)))
col[grepl("HBC", rownames(DMR.CGI))] <- "forestgreen"
op <- par(mai = rep(0.5, 4))
heatmap.2(DMR.CGI, col = redblue(256), RowSideColors = col, density.info = "none", 
          trace = "none", Rowv = TRUE, Colv = TRUE, labCol = FALSE, labRow = FALSE, 
          dendrogram = "row", margins = c(1, 5))
legend("topright", c("ALL", "HBC"), col = c("darkgoldenrod1", "forestgreen"), 
       pch = 15)
```

To plot location of differential methylated probes along each chromosome
```{r fig.width=7, fig.height=6}
# get the length of chromosome 1-22 and X
chrlen <- unlist(as.list(IlluminaHumanMethylation450kCHRLENGTHS)[c(as.character(1:22), 
    "X")])
chrlen <- data.frame(chr = factor(names(chrlen)), length = chrlen)
chr <- IlluminaHumanMethylation450kCHR  # get the chromosome of each probe
# get the probe identifiers that are mapped to chromosome
chr <- unlist(as.list(chr[mappedkeys(chr)]))
# get chromosome coordinate of each probe
coord <- IlluminaHumanMethylation450kCPGCOORDINATE
# get the probe identifiers that are mapped to coordinate
coord <- unlist(as.list(coord[mappedkeys(coord)]))
coord <- data.frame(chr = chr[intersect(names(chr), names(coord))], coord = coord[intersect(names(chr), 
    names(coord))])
# coordinates of probes in DM CGIs
coordDMRprobe <- droplevels(na.omit(coord[cginame[cginame$cginame %in% rownames(DMR), 
    ]$Probe_ID, ]))
(coord.plot <- ggplot(data = coordDMRprobe) + geom_linerange(aes(factor(chr, 
    levels = c("X", as.character(22:1))), ymin = 0, ymax = length), data = chrlen, 
    alpha = 0.5) + geom_point(aes(x = factor(chr, levels = c("X", as.character(22:1))), 
    y = coord), position = position_jitter(width = 0.03), na.rm = T) + ggtitle("DMR positions on chromosomes") + 
    ylab("Position of DMRs") + xlab("chr") + coord_flip() + theme_bw())
```


